Inner loop optimizations in mapping single-threaded programs to hardware

### 2014-11-05 Documents/Publications/loopPipelining/paper.tex
Revived loop-pipelining paper for arxiv.

\author{Author's name Suppressed for Blind Review}
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]

@@

  In the context of mapping high-level algorithms to hardware,
  we consider the basic problem of generating an efficient hardware 
  implementation of a single threaded program, in particular, that
  of an inner loop.  
  We describe a control-flow mechanism through
  which multiple iterations of an arbitrary inner loop can be made simultaneously
  active in the generated hardware, thus providing dynamic loop-pipelining
  capability in hardware.   
  We study the impact of this loop-pipelining scheme in conjunction with
  source-level loop-unrolling.
  In particular, we consider some common loop kernels occurring in fast-fourier transform,
  matrix multiplication and inner product computations, with the resulting
  hardware descriptions synthesized to an FPGA target.  We observe
  that the use of dynamic loop-pipelining mechanism alone typically results in
  a significant improvements in the performance of the hardware.  If the loop is
  statically unrolled and if loop-pipelining is applied to the unrolled
  program, then the performance improvement is still substantial.  When
  both optimizations are used in conjunction, the improvement in performance
  ranges from 6X to 20X (in terms of number of clock cycles needed for the computation)
  across the loop kernels that we have studied.
  These optimizations do have a hardware overhead, but,
  in spite of this, we observe that the joint use of these 
  loop optimizations not only improves performance, but also the 
  performance/cost ratio of the resulting hardware.

@@

%% The importance and problem of loops.
A high-level-synthesis (HLS) compiler takes an algorithm described in
a high-level programming language (such as C, for example) and produces a circuit
description (a register-transfer-level description in a hardware
description language such as VHDL, for example).
High-level synthesis systems have been available for some time
now (for example, \cite{pegasus-cash}, \cite{spark-vlsi-paper} and
commercial tools such as Catapult-C\footnote{\url{http://www.mentor.com/products/esl/high_level_synthesis/}} from Mentor Graphics
C-to-Silicon\footnote{\url{http://www.cadence.com/products/sd/silicon_compiler/}} from Cadence,
Synopsys HLS\footnote{\url{http://www.synopsys.com/Tools/SLD/HLS/}} from Synopsys).

The main attraction of such a compiler is a substantial reduction in 
the design and verification effort.  The
main concern is the quality of the hardware that is produced.
For the same starting point, the quality of the hardware produced may be objectively
measured in terms of factors such as area, energy-consumption and performance (in the
sense of time required for completion, or in terms of the rate at which the hardware
finishes the computational tasks).  An {\em internal measure} of the quality of the
hardware produced is the utilization factor of the major resources in the hardware (that is,
the utilization factor of expensive resources).

In our work, we have developed a set of compiler tools which take a C-description of
an algorithm (with the only important restrictions being the requirement that
cycles are not present in the call graph of the source program, and the lack of 
support for {\bf function} pointers, normal pointers are supported), and produces
a VHDL description which is provably equivalent to the original program \cite{SameerPhD}.
Results on the application of this compiler to single-threaded programs show
that the circuits (when synthesized to VLSI layout) are substantially more
energy efficient than a processor, but that the performance of the processor
is usually better than that of the synthesized circuit \cite{ahirDsd2010}.
If multi-threaded source programs are used as a starting point, the circuits produced 
are quite competitive with hand-designed RTL \cite{ahirUsenix2012}.
Quantitatively, the results in \cite{ahirDsd2010} and \cite{ahirUsenix2012} indicate
that the gap between the synthesized hardware and modern
pipelined processors (for single-threaded
programs) and between the synthesized hardware and hand-crafted hardware (for
multi-threaded programs) is of the order of 2-3X.  That is, the performance
of the synthesized hardware needs to improve by 2-3X in order to close
the respective performance gaps.

%% How compilers have handled this problem.
In this paper, we consider the problem of improving the performance
of hardware generated from single threaded programs; in
particular, the problem of mapping loops to hardware.  It
is well known that most compute intensive programs spend a large fraction
of their time in inner loops.  Thus, the optimal implementation
of such loops is of primary importance, whether the target is
a processor or hardware.  Such an improvement is essential if
synthesized hardware is to be performance competitive with
high performance processors or with hand-crafted hardware.

@@

The observations indicate the following:
\begin{itemize}
\item The performance improvement with loop-pipelining applied alone is in the
2X-3X range. 
\item The performance improvement with loop-unrolling alone is in the 
2X range, and with aggressive unrolling (as was tried in the matrix multiplication
case), the improvement is as high as 10X.
\item The combination of loop-pipelining and loop-unrolling leads
to a performance improvement which is at least as high as the
product of the improvements due to the individual optimizations.  
For matrix multiplication, this improvement is as large as 20X.
\item The hardware overheads for implementing these optimizations are considerable,
but the cost-to-performance ratio improves substantially in all cases.
\end{itemize}
The results indicate that loop-unrolling combined with dynamic loop-pipelining
can close the performance gap noted above.

For the subsequent discussion, we assume that the inner
loop which is being optimized consists of a single
basic block, that is, there are no branching instructions
in the loop body (no jumps, if constructs, switch constructs etc.).
If such constructs are present, we assume that they have been
eliminated using the mechanism of guarded (or predicated) execution.

@@

We give a brief description of our compiler flow.  The
details are not relevant to this paper, and the interested
reader can find them in \cite{ahirDsd2010}.

@@

The compiler flow has been characterized over a wide variety
of applications \cite{ahirDsd2010}.   The XXXX\footnote{name suppressed for blind review.} tool-chain \cite{ahirUsenix2012} 
uses this compiler flow to map modular network appliances written in C++ to
an FPGA platform, and achieved about half the
performance possible using equivalent hand designed hardware.  
Thus,  the flow can be applied to ``real'' software
to produce competitive results.  The optimizations indicated in this
paper, should in principle, make the flow even more competitive.

%% compiler techniques.

%% hls techniques to handle loops?

@@

\subsection{Control-path}
%%
%%  control path and data-flow
%%

@@

\section{A control-flow mechanism for loop-pipelining}

%%
%% reenable rules.
%%
%% loop-terminator.
%%
%% advantages of dynamic pipelining: any loop
%% can be handled..
%%
Suppose that we want to modify the control-path in order to
permit the second (and maybe third etc.) iteration of a loop
to begin while the first iteration is still in progress.  
Consider the example of the dot product.  The original
loop was
\begin{verbatim}
float a[1024], b[1024];
float dotp = 0.0;
for(i=0; i < 1024; i++)
{
   dotp += a[i]*b[i];
}
\end{verbatim}
The {\em fully unrolled} version of this loop would be
\begin{verbatim}
float a[1024], b[1024];
float dotp = 0.0;
dotp += a[0]*b[0];
dotp += a[1]*b[1];
dotp += a[2]*b[2];
...
dotp += a[1023]*b[1023];
\end{verbatim}
Let $A$ denote the $*$ operation and let $B$ denote the $+$
operation, $L_a,\ L_b$ denote the loads from $a$ and $b$ respectively.  
In principle, all the loads can occur simultaneously, and all the multiplies
can happen simultaneously once the loads complete.  The adds
would need to be ordered because of the multiply-accumulate nature
of the code as it is written.  Any ordering of these operations which
satisfies these dependencies will be termed a {\em fully-unrolled consistent}
ordering.

@@

It is easy to confirm that these additional dependencies ensure
that the loop execution subject to these dependencies is a
fully-unrolled consistent ordering.  
The modified control path is shown in Figure \ref{fig:pipelinedCP}.
The loop-terminator element has three inputs:  a loop-taken transition,
a loop-not-taken transition and a loop-body-exit transition.  The loop-taken/not-taken
pair indicates whether a new iteration is to be started or whether the 
loop has terminated.  The loop-body-exit
transition indicates that the body of the loop has finished executing an
iteration.  The loop-terminator initiates a new iteration as long as the
number of active iterations is within a specified limit $M$ (usually, we keep
this limit to $M=8$).  Thus, all the places in the modified control path
in Figure \ref{fig:pipelinedCP} now must have a capacity of $M$ and the
cost of implementing each place in the control path goes up by a
factor of $\log M$.  This is the major additional cost incurred by
the pipelining mechanism.
\begin{figure}[ht]
  \centering
  \includegraphics[width=7cm]{pipelinedCP.eps}
  \caption{Modified control-path with loop-pipeline dependencies (dotted lines).}
  \label{fig:pipelinedCP}
\end{figure}

@@

We considered three examples, the dot product, the fast-fourier-transform,
and matrix multiplications, each with critical inner loops.  Four 
configurations were tested in each case: the basic code, the unrolled code,
the basic code with loop-pipelining and the unrolled code with loop-pipelining.
In each case, the generated VHDL code was synthesized to a Xilinx Virtex-6
FPGA and synthesis results were used to estimate the resource usage,
the clock frequency and the number of cycles required by the inner
loop.

@@

@@

The data obtained from three common and important inner loop
kernels is encouraging.  In all three cases, both loop-pipelining
and loop-unrolling lead to substantial performance gains.  Further,
using both optimizations together results in multiplicative gains
and in call cases, leads to hardware which is substantially
faster and more efficient (in terms of the performance/cost ratio).
The performance gain is lower if there is a bottleneck in the
algorithm itself, such as in the FFT case, in which accesses to
the in-place array reduce the performance gains seen due to
the loop optimizations.

Loop-unrolling is a common parallelization optimization
in compilers that target pipelined processors.  Our data indicates
that such unrolling offers significant benefits (up to an 
order of magnitude) in an algorithm-to-hardware translation.
Further, the use of hardware based dynamic loop-pipelining techniques 
seems to offer a further boost in performance in the synthesis
of hardware from single-threaded programs.  
The performance boost provided by the dynamic loop-pipelining
in hardware seems to indicate that its use,
especially in conjunction with aggressive loop unrolling can offer 
a substantial reduction in the gap between 
the quality of automatically generated hardware and hand crafted 
hardware implementations of the same algorithm.  This needs
to be investigated further.

>>>>

\author{Madhav P. Desai\\ Department of Electrical Engineering\\ IIT Bombay, Mumbai, India\\ madhav@ee.iitb.ac.in}
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]

@@

  In the context of mapping high-level algorithms to hardware,
  we consider the basic problem of generating an efficient hardware 
  implementation of a single threaded program, in particular, that
  of an inner loop.  
  We describe a control-flow mechanism which provides dynamic loop-pipelining
  capability in hardware, so that multiple iterations of an arbitrary inner loop 
  can be made simultaneously active in the generated hardware, 
  We study the impact of this loop-pipelining scheme in conjunction with
  source-level loop-unrolling.  
  In particular, we apply this technique to some common loop kernels: regular kernels
  such as the  fast-fourier transform and matrix multiplication, as well
  as an example of an inner loop whose body has branching.
  The resulting hardware descriptions are synthesized to an FPGA target,
  and then characterized for performance and resource utilization.  
  We observe
  that the use of dynamic loop-pipelining mechanism alone typically results in
  a significant improvements in the performance of the hardware.  
  If the loop is
  statically unrolled and if loop-pipelining is applied to the unrolled
  program, then the performance improvement is still substantial.  
  When dynamic loop pipelining is used in conjunction with static loop unrolling, the improvement in performance
  ranges from 6X to 20X (in terms of number of clock cycles needed for the computation)
  across the loop kernels that we have studied.
  These optimizations do have a hardware overhead, but,
  in spite of this, we observe that the joint use of these 
  loop optimizations not only improves performance, but also the 
  performance/cost ratio of the resulting hardware.

@@

%% How compilers have handled this problem.
We consider the problem of improving the performance
of hardware generated from single threaded programs; in
particular, the important problem of mapping loops to hardware.  It
is well known that most compute intensive programs spend a large fraction
of their time in inner loops.  Thus, the optimal implementation
of such loops is of primary importance, whether the target is
a processor or hardware.  Such an improvement is essential if
synthesized hardware is to be performance competitive with
high performance processors or with hand-crafted hardware.

%% The importance and problem of loops.
% A high-level-synthesis (HLS) compiler takes an algorithm described in
% a high-level programming language (such as C, for example) and produces a circuit
% description (a register-transfer-level description in a hardware
% description language such as VHDL, for example).
% High-level synthesis systems have been available for some time
% now (for example, \cite{pegasus-cash}, \cite{spark-vlsi-paper} and
% commercial tools such as Catapult-C\footnote{\url{http://www.mentor.com/products/esl/high_level_synthesis/}} from Mentor Graphics
% C-to-Silicon\footnote{\url{http://www.cadence.com/products/sd/silicon_compiler/}} from Cadence,
% Synopsys HLS\footnote{\url{http://www.synopsys.com/Tools/SLD/HLS/}} from Synopsys).

% The main attraction of such a compiler is a substantial reduction in 
% the design and verification effort.  The
% main concern is the quality of the hardware that is produced.
% For the same starting point, the quality of the hardware produced may be objectively
% measured in terms of factors such as area, energy-consumption and performance (in the
% sense of time required for completion, or in terms of the rate at which the hardware
% finishes the computational tasks).  An {\em internal measure} of the quality of the
% hardware produced is the utilization factor of the major resources in the hardware (that is,
% the utilization factor of expensive resources).

% In our work, we have developed a set of compiler tools which take a C-description of
% an algorithm (with the only important restrictions being the requirement that
% cycles are not present in the call graph of the source program, and the lack of 
% support for {\bf function} pointers, normal pointers are supported), and produces
% a VHDL description which is provably equivalent to the original program \cite{SameerPhD}.
% Results on the application of this compiler to single-threaded programs show
% that the circuits (when synthesized to VLSI layout) are substantially more
% energy efficient than a processor, but that the performance of the processor
% is usually better than that of the synthesized circuit \cite{ahirDsd2010}.
% If multi-threaded source programs are used as a starting point, the circuits produced 
% are quite competitive with hand-designed RTL \cite{ahirUsenix2012}.
% Quantitatively, the results in \cite{ahirDsd2010} and \cite{ahirUsenix2012} indicate
% that the gap between the synthesized hardware and modern
% pipelined processors (for single-threaded
% programs) and between the synthesized hardware and hand-crafted hardware (for
% multi-threaded programs) is of the order of 2-3X.  That is, the performance
% of the synthesized hardware needs to improve by 2-3X in order to close
% the respective performance gaps.

@@

The observations indicate the following:
\begin{itemize}
\item The performance improvement with loop-pipelining applied alone is in the
2X-8X range.   This improvement is observed both in the case of an inner
loop whose body is a single basic block, as well as in the case when the
inner loop body has branches.
\item The performance improvement with loop-unrolling alone is in the 
2X range, and with aggressive unrolling (as was tried in the matrix multiplication
case), the improvement is as high as 10X.
\item The combination of loop-pipelining and loop-unrolling leads
to a performance improvement which is at least as high as the
product of the improvements due to the individual optimizations.  
For matrix multiplication, this improvement is as large as 20X.
\item The hardware overheads for implementing these optimizations are considerable,
but the cost-to-performance ratio improves substantially in all cases.
\end{itemize}
The results indicate that loop-unrolling combined with dynamic loop-pipelining
can close the performance gap noted above.

@@

The dynamic pipelining mechanism described in this paper
is implemented in a compiler flow which takes 
a C program and produces an equivalent VHDL description.
We give a brief description of this compiler flow.  The
details are not relevant to this paper, and the interested
reader can find them in \cite{ahirDsd2010}.

@@

The compiler flow has been characterized over a wide variety
of applications \cite{ahirDsd2010}, \cite{ahirUsenix2012}.

@@

\subsection{Control-path}

%%
%%  control path and data-flow
%%

@@

\section{A control-flow mechanism for dynamic loop-pipelining}

For the subsequent discussion, we assume that the inner
loop which is being optimized consists of a single
basic block, that is, there are no branching instructions
in the loop body (no jumps, if constructs, switch constructs etc.).
If such constructs are present, these are first eliminated
using the mechanism of guarded (that is, predicated) execution.

%%
%% reenable rules.
%%
%% loop-terminator.
%%
%% advantages of dynamic pipelining: any loop
%% can be handled..
%%
Suppose that we want to modify the control-path in order to
permit the second (and maybe third etc.) iteration of a loop
to begin while the first iteration is still in progress.  
Consider the example of the dot product.  The original
loop was
\begin{verbatim}
float a[1024], b[1024];
float dotp = 0.0;
for(i=0; i < 1024; i++)
{
   dotp += a[i]*b[i];
}
\end{verbatim}
The {\em fully unrolled} version of this loop would be
\begin{verbatim}
float a[1024], b[1024];
float dotp = 0.0;
dotp += a[0]*b[0];
dotp += a[1]*b[1];
dotp += a[2]*b[2];
...
dotp += a[1023]*b[1023];
\end{verbatim}
Let $A$ denote the $*$ operation and let $B$ denote the $+$
operation, $L_a,\ L_b$ denote the loads from $a$ and $b$ respectively.  
In principle, all the loads can occur simultaneously, and all the multiplies
can happen simultaneously once the loads complete.  The adds
would need to be ordered because of the multiply-accumulate nature
of the code as it is written.  Any ordering of these operations which
satisfies these dependencies will be termed a {\em loop-consistent}
ordering.

@@

It is easy to confirm that these additional dependencies ensure
that the loop execution subject to these dependencies is a
loop-consistent ordering.  
The modified control path is shown in Figure \ref{fig:pipelinedCP}.
The loop-terminator element has three inputs:  a loop-taken transition,
a loop-not-taken transition and a loop-body-exit transition.  The loop-taken/not-taken
pair indicates whether a new iteration is to be started or whether the 
loop has terminated.  The loop-body-exit
transition indicates that the body of the loop has finished executing an
iteration.  The loop-terminator initiates a new iteration as long as the
number of active iterations is within a specified limit $M$ (usually, we keep
this limit to $M=8$).  Thus, all the places in the modified control path
in Figure \ref{fig:pipelinedCP} now must have a capacity of $M$ and the
cost of implementing each place in the control path goes up by a
factor of $\log M$.  This is the major additional cost incurred by
the pipelining mechanism.
\begin{figure}[ht]
  \centering
  \includegraphics[width=7cm]{pipelinedCP.eps}
  \caption{Modified control-path with loop-pipeline dependencies (dotted lines).}
  \label{fig:pipelinedCP}
\end{figure}

@@

We considered four examples:
\begin{itemize}
\item Three examples where the loop body was a single basic block: the dot product, the fast-fourier-transform,
and matrix multiplications, each with critical inner loops.  Four 
configurations were tested in each case: the basic code, the unrolled code,
the basic code with loop-pipelining and the unrolled code with loop-pipelining.
In each case, the generated VHDL code was synthesized to a Xilinx Virtex-6
FPGA and synthesis results were used to estimate the resource usage,
the clock frequency and the number of cycles required by the inner
loop.
\item One example where the loop body exhibits branching: a stream processor kernel which
operates on a stream of numbers and performs operations on the stream depending on
an op-code stream.  This example illustrates that the dynamic loop-pipelining mechanism
is effective in complex loop bodies as well.
\end{itemize}

@@

\section{A stream processor}

The following loop was used to test a situation in
which the loop body has branching.  
\begin{verbatim}
while(1)
{
  float x = read_float32("x_pipe");
  float y = read_float32("y_pipe");
  uint8_t op_code = read_uint8("op_pipe");

  float result = 0;
  if(op_code == 0)
    result = x*y;
  else if(op_code == 1)
    result = x+y;
  else if(op_code == 2)
    result = (x*x) - (y*y);
  else if(op_code == 3)
    result = (x + y) * (x + y);
  else
    result = 0;
  write_float32("z_pipe",result);
}
\end{verbatim}
In this loop, $x,y$ and $op\_code$ are read from
three input streams.  Depending on the value of
$op\_code$, a result is calculated and written out
to an output stream.

In order to pipeline this loop, the conditional
statements are first eliminated using guards.   This
is done by calculating the predicates 
\begin{verbatim}
(op_code == 0)
(op_code == 1)
(op_code == 2)
(op_code == 3)
\end{verbatim}
and using these predicates to guard the execution of
the statements which depend on these conditions.  This
is done automatically in our compiler.

The observations are shown in Table \ref{table:streamProcessor}.
The time reported is that needed to process 16 elements from
the streams (that is, to complete 16 iterations of the loop).
\begin{table}[htb]
  \centering
  \caption{Stream processor inner loop observations: pipelined versus non-pipelined}
  \label{table:streamProcessor}
  \renewcommand\arraystretch{1.2}
  \setlength{\tabcolsep}{1ex}
  \begin{tabular}{c|c|c}
  \hline
  & plain & pipelined \\
  \hline

Cycles   &    2913  &   334  \\
LUTs     &    4603  &  12501  \\
FFs      &    4248  &  8808  \\
Freq.(MHz)  & 184.9  &   103.9 \\
Norm. Perf. & 1      &  4.86 \\
Norm. Perf/Cost & 1  &  2.01  \\
\hline
  \end{tabular}
\end{table}

We observe a 4X improvement in performance and a 2X improvement
in the performance/cost ratio.

@@

The data obtained from four inner loop
kernels is encouraging. The first three  were examples in
which the inner loop body consisted of a single basic 
block. In these cases, both loop-pipelining
and loop-unrolling lead to substantial performance gains.  Further,
using both optimizations together results in multiplicative gains
and in call cases, leads to hardware which is substantially
faster and more efficient (in terms of the performance/cost ratio).
The performance gain is lower if there is a bottleneck in the
algorithm itself, such as in the FFT case, in which accesses to
the in-place array reduce the performance gains seen due to
the loop optimizations.
In the fourth inner loop kernel, we considered a loop body
which had branching.  In this case, considerable performance
and performance/cost gains were observed when loop pipelining
was enabled.

Thus, the use of hardware based dynamic loop-pipelining techniques 
offers a significant boost in performance in 
hardware synthesized from single-threaded programs.  
The performance boost provided by the dynamic loop-pipelining
in hardware seems to indicate that its use,
especially in conjunction with aggressive loop unrolling can offer 
a substantial reduction in the gap between 
the quality of automatically generated hardware and hand crafted 
hardware implementations of the same algorithm.  This needs
to be investigated further.
